An investigation into using scagnostics interesting projections of high-dimensional data

### 2018-09-28 scagnostics.Rmd
added some outlines for sections

---
title: Using Scagnostics to find Interesting Projections of Multivariate Data

@@

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE, 
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = TRUE, external = TRUE, dev = "pdf",
  fig.height = 5, fig.width = 8, out.width = "\\textwidth"
)
```

@@

The term "projection pursuit" was coined by @ff74 to describe a procedure for searching high-(say $p$)dimensional data for "interesting" low- dimensional projections ($k=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $k$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $k$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not necessarily found by methods such as principal component analysis.

A large number of projection pursuit indices have been developed [@f87] [@CBC92] [@Lee05] [@AHC02] [@CEM:CEM2568] [@JS87] [@5508437]. Many of these are developed to detect departure from multivariate normal, which includes clusters or outliers, or nonlinearity. Some have been developed to detect separations between known groups. However, if one wants to find complicated codependence between parameters there are no options.
On the other hand, for the related question of the selection of pair-wise scatter plots in datasets with large numbers of parameters, a number of different methods have been developed, from detection of linear relations by correlation coefficients, to more complex measures aiming to capture generally the information in a given view.
Here we want to extend the list of projection pursuit indices with the aim of detecting non-trivial dependence between parameter combinations, i.e. we want to find complicated functional dependence as well as association not described by a function (e.g. an ellipse or non-coexistence).
Our motivation is based on potential physics applications aiding the interpretation of model fits to experimental results. Let us consider a physical model with a set of $n$ free parameters that cannot be measured directly but should be determined from fitting a set of $p$ experimental observations for which predictions can be calculated when fixing all $n$ parameters. Note here that while we often have analytic expressions for the predictions, this is not always the case and we may have to rely on numeric calculations. Moreover, a single prediction can be a complicated function of all free parameters. In addition we often deal with large number of observations, $p \sim 100-1000$ as well as sizable parameter spaces $n \sim 10$. The results of a fit are generally interpreted using visualisations selected based on physics intuition and previous knowledge, but the above stated observations suggest that some correlations clearly dictated by the data may be missed, as we are typically only studying selected views of the results. The idea of finding special views in scatter plots fits naturally with the concept of using posterior samples obtained in a Bayesian fit.
The idea is that by finding special views of the posterior samples we gain insight into the physics relevant to accurately describe the experimental data. Apart from improved intuition, finding special regions in parameter space is important as it allows us to further scrutinize the model. There are many examples of this, for example in particle physics detailed studies have been performed e.g. in the alingment limit of the 2HDM or in coannihilation regions in models of dark matter.

A related question is the selection of pair-wise scatter plots in datasets with large numbers of parameters, for which different methods have been developed. One approch are scatterplot diagnostics, "scagnostics", developed by @scag, and explored in @WW08. \footnote{Scagnostics are available in two R packages @LWscagR and @HWscagR. This work is built upon the scagnostics package by @HWscagR because the underlying base is built using C, works across platforms.} They are calculated from a set of elementary graph-theoretic building blocks: the convex hull, the alpha hull and the minimal spanning tree (MST). These are used to define eight measures: Outlying, Skewed, Sparse, Clumpy, Striated, Convex, Skinny and Stringy. While in principle each measure could be used on its own to define a ranking, the original idea suggested considering pair-wise scatter plots of the scagnostics measures for identification of "characteristic" and "unusual" 2-d views in the original parameter space. Alternative methods suggested by (Katrins Thesis or maybe her R package?) suggest using parallel coordinates, glyphs or a so-called scagramm (based on the idea of corrgrams) that shouold be coupled to interactive brushing for selection of pair-wise scatter plots to be displayed.
(moved scagnostics definitions to appendix, not sure if needed at all.)
(Issues with scagnostics already for selection of pairs have been described in Katrins Thesis, should have short summary of that here?)

Given the various shortcomings of the scagnostics approach we consider several other measures that can directly be interpreted as ranking of information contained in a given 2-d view.
Two such additional measures have been defined in @mbgraphic. The first is based on smoothing splines, and is defined such that it will take maximum values when the data distribution can be described by a functional form. For a scatterplot with variable X and Y we calculate it as
\begin{equation}
splines_{2d} = max(1-\frac{Var(res_{X\sim Y})}{Var(X)}, 1-\frac{Var(res_{Y\sim X})}{Var(Y)})
\end{equation}
where $Var(.)$ is the variance and $res_{X\sim Y}$ are the residuals in the spline model where $X$ is considered a function of $Y$.
The second new measure introduced in @mbgraphic is aiming to detect also non-functional relations and is based on the idea of distance correlation [@szekely2007]. (need to add definition)
Finally we will also consider the maximal and total information coefficient (MIC and TIC) defined in @Reshef1518 which are based on mutual information found when imposing grids on pair-wise variable plots. These functions are availabel in R via the interface implemented in the minerva package [@minerva]. 
(read paper and add more detail. how are they similar/different compared to scagnostics and distance correlation? the way that binning is being built into it may make difference in how MIC/TIC may be smoother?)

A projection pursuit index, a function of all possible projections of the data, invariably has many "hills and valleys" and "knife-edge ridges" because of the varying shapes in the underlying density of observations from one projection to the next. From an exploratory data analysis perpective it is interesting to combine numerical optimisation with visualisation to watch the structure of the data moving into a maximum, to jump away from this projection and follow the optimisation again, invariably moving from local maxima to local maxima, and perhaps even a global maximum. Each of these maxima can reveal different information about the data. Projection pursuit optimisation combined with geodesic interpolations between projections provided by a tour [@As85], is called a projection pursuit guided tour [@CBCH94]. Software to run a projection pursuit guided tour is available in the R package, "tourr" [@tourr].
(Motivation strongly relates to intuition, i.e. guided tour, local tour and manual tour may be useful to understand structure in the multivariate context.)
(Even just for the optimisation part the guided tour implementation is useful.)
(Traditional projection pursuit indices are fast to compute and therefore the optimisation would be done in "real time" while displaying the tour. For the measures considered here this generally does not apply, as a consequence we "disentangle" the two into subsequent steps.)

@@

@@

@@

((maybe better to start with something like this, though maybe too obvious?))
The various measures defined in the introduciton are designed to be computed for all pairs of variables, followed either by selection based on comparing scagnostics scores between all pairs, or by an absolute ranking from most to least interesting (e.g. when using TIC).
Here we modify the usage as follows. The tour implementation is used to generate projections of the multivariate dataset down to two dimensions, i.e. the projected data can be represented as a scatter plot. In this projection the data is represented by two parameters that are defined as combinations of the input parameters, and this parametrisation is passed into the index function calculation to obtain a score for the selected view. Using the guided tour implementation the index functions are optimized such that we find the generalised 2-d parametrisation yielding the most intersting view of the data.

To use an index function for projection pursuit requires that the measure be relatively smooth over the space of all 2D projections, and we will first investigate the smoothness over randomly selected projections. An additional consideration is the computing time, ideally fast computation is preferred as the guided tour can be viewed as the index is being optimized. However, a compuationally more costly index may be preferred for better detection properties, and one can consider recording the guided tour for viewing after the optimisation has converged. Finally we note that another important consideration is how the index function changes as the tour moves into the "optimal" view, this will however depend mainly on the distribution rather than the selected index function. (we will see this in the examples)

@@

1. p points on an n-dimensional sphere randomly selected using the geozoo package, as an example of dataset without special views
2. p points where n-1 points are independently drawn from a uniform distribution between [-1,1], and enforcing $x_n^2 + x_{n-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparant in the distribution 
3. p points where n-1 points are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_n = \sin(x_{n-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function

@@

To get an overview of the dataset we show views of the first two parameters showing the independent distribution and the last two parameters showing the special view. The enforced relationships will also affect the 1-d densities as can be seen from \ref{fig:dataPlotsDensity}, which means that views including sizable fractios of V6 (and V5 for the Pipe distribution) will be distinct from uninformed views of distributions 2 and 3. We select two samples for each distribution, with 100 and 1000 sampled points respectively.

```{r dataPlotsV1V2, fig.width=6, fig.height=7, fig.cap="Data views of V1 vs V2."}
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsV5V6, fig.width=6, fig.height=7, fig.cap="Data views of V5 vs V6."}
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V5, V6)) +
    geom_point() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

@@

We next compare the various index values for three views: V1 vs V2, V5 vs V6 and V1 vs V6 (capturing differences from the different distribution in V6). For the Sphere distribution by construction we see only random fluctuations between the different views, which can be sizable in particular for small datasets. Considering next the Pipe distribution, we see that skinny, stringy and dcor2d are increased for the special view, but in particular MIC and TIC index are much larger and may be used to detect this type of structure in the dataset. Moreover, minimising the convex index (or maximising 1-convex) should also allow us to detect the special view. Finally the Sine distribution shows that several of the indices are maximised in the special view, in particular also the splines2d index designed for the detection of fuctional dependence.

```{r indexTable, echo=FALSE}
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
indexOverview <- tibble(convex=numeric(),
               clumpy=numeric(),
               skinny=numeric(),
               stringy=numeric(),
               dcor2d=numeric(),
               splines2d=numeric(),
               MIC=numeric(),
               TIC=numeric(),
               name=character(),
               vars=character())
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  ds <- as.tibble(rescale(ds))
  for(vars in list(c("V1","V2"),c("V5","V6"),c("V1","V6"))){
    dprj <- as.matrix(select_(ds, vars[1], vars[2]))
    scagRes <- scagnostics(dprj)
    dcorRes <- dcor2d(dprj[,1], dprj[,2])
    splineRes <- splines2d(dprj[,1], dprj[,2])
    mineRes <- mine(dprj[,1], dprj[,2])
    indexOverview <- add_row(indexOverview, convex=scagRes["Convex"],
                    clumpy=scagRes["Clumpy"], skinny=scagRes["Skinny"],
                    stringy=scagRes["Stringy"], dcor2d=dcorRes,
                    splines2d=splineRes, MIC=mineRes$MIC,
                    TIC=mineRes$TIC, name=pLabels[i], vars=paste0(vars[1],"-",vars[2]))
  }
  i <- i+1
}
knitr::kable(indexOverview ,  caption = "Summary of index values.", digits=2) 
```

@@

```{r getProj}
getProj <- function(df, tPath, nameStr, size){
  sc <- tibble(
    holes=numeric(),
    cmass=numeric(),
    convex=numeric(),
    clumpy=numeric(),
    skinny=numeric(),
    stringy=numeric(),
    dcor2d=numeric(),
    splines2d=numeric(),
    MIC=numeric(),
    TIC1=numeric(),
    t=numeric(),
    name=character(),
    size=numeric())
  n <- length(tPath)
  df <- rescale(df)
  for (i in 1:n) {
    dprj <- df %*% tPath[[i]]
    scagRes <- scagnostics(dprj)
    dcorRes <- dcor2d(dprj[,1], dprj[,2])
    splineRes <- splines2d(dprj[,1], dprj[,2])
    mineRes <- mine(dprj[,1], dprj[,2])
    holesRes <- holes(dprj)
    cmassRes <- cmass(dprj)
    sc <- add_row(sc, holes=holesRes, cmass=cmassRes,
                  convex=scagRes["Convex"],
                  clumpy=scagRes["Clumpy"], skinny=scagRes["Skinny"],
                  stringy=scagRes["Stringy"], dcor2d=dcorRes,
                  splines2d=splineRes, MIC=mineRes$MIC,
                  TIC1=mineRes$TIC, t=i, name=nameStr, size=size)
  }
  maxTIC <- max(sc$TIC1)
  sc <- sc %>%
    mutate(TIC = TIC1/maxTIC) %>%
    select(-TIC1)

@@

Sharp jumps can be observed in all measures except the distance correlation measure when considering small datasets of only 100 points. Generally we observe smoother behaviour for larger datasets, notably all considered scagnostics measures show only small variation around a flat baseline. Interestingly all other indices indicate preferred viewing directions for dataset 3 when studying large samples, but not for dataset 2 which contains a special view in the same direction. We therefore show 2-d view defined by the tour step $t=58$, with projection matrix
```{r t58matrix, echo=FALSE}
knitr::kable(t1full[[58]] ,  caption = "Matrix representation of projection 58.")
```
i.e. while the last two parameters enter significantly, they will not dominate the view and we do not expect to fully observe the built in features.
```{r plot58, fig.width=6, fig.height=7, fig.cap="View of all datasets defined by at tour step 58."}
p58 <- t1full[[58]]
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  dm <- rescale(ds)
  pData <- as.tibble(dm %*% p58)
  pC <- ggplot(pData, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]],
             pL[[3]], pL[[4]],
             pL[[5]], pL[[6]],
             ncol=2)

```
We see from Fig. ((how to add cross references?)) that the built in special views cannot be seen in this projection, but confirm that the sine 1000 dataset presents some correlation in this view.

## Transition
We repeat the same exercise, but now we are interested in how the index values change when going from the view in V1 vs V2 to the view in V5 vs V6.
```{r plannedtourpath}
m1 <- matrix(c(1,0,0,0,0,0,0,1,0,0,0,0),ncol=2)
m2 <- matrix(c(0,0,0,0,1,0,0,0,0,0,0,1),ncol=2)
#this is silly but seems that first two entries are being ignored, so need some fake entries
m3 <- matrix(c(1,0,0,0,1,0,0,1,0,0,0,1),ncol=2)
m4 <- matrix(c(1,1,1,0,0,0,0,1,1,0,0,0),ncol=2)
t2 <- save_history(sin100,tour_path=planned_tour(list(m3,m4,m1,m2)))
t2full <- as.list(interpolate(t2))
fullResPlanned <- getProj(sin100, t2full, "sin", 100) %>%
  rbind(getProj(sin1000, t2full, "sin", 1000)) %>%
  rbind(getProj(sphere100, t2full, "sphere", 100)) %>%
  rbind(getProj(sphere1000, t2full, "sphere", 1000)) %>%
  rbind(getProj(pipe100, t2full, "pipe", 100)) %>%
  rbind(getProj(pipe1000, t2full, "pipe", 1000))

@@

@@

We can see a pretty clear and smooth increase in dcor2d and splines2d as well as MIC and TIC index when approaching the special view of the Sine distribution, which should therefore be efficient to optimise. On the other hand while we also see smooth increase in MIC and TIC when approaching the special view of the Pipe distribution, the increase only becomes apparant much closer to the final projection, making this structure more challenging to detect with a guided tour.
Note that the various indices appear less noisy than for the grand tour shown before, but should note that the tour path here is shorter, and the y-axis scale is different (typically reaching higher index values in this example, whereas smaller fluctuations are more amplfied in the previous views).
(What does this mean for optimisation? How can I detect such "smaller scale increase"? Maybe try "search_better" instead of search_geodesic? increase alpha and decrease cooling? increase max.tries?)
Can force it when using search_better with large alpha, cooling=1, max.tries=1000, but not ideal..
Difficulties also from "fake" structure from the sampling boundaries?
Possible issue with search_geodesic is that the search window is fixed in the code..

@@

Here we also see how the existing indices in the tourr package are inadequat for the detection of these type of dependence, the holes index taking maximum values before the final view and decreasing when moving into the special view. On the other hand, surprisingly the cmass index increases when moving into the special view, but the intermediate minimum makes it unsuitable as a projection pursuit index here. QUESTION: how come they are all almost the same for the various distributions? Especially pipe and sphere seem to give almost identical behaviour, can it be understood from the index definition?

>>>>

---
title: An investigation into using scagnostics interesting projections of high-dimensional data

@@

```{r initial, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE, 
  fig.path = 'figure/', cache.path = 'cache/', fig.align = 'center', 
  fig.show = 'hold', cache = FALSE, external = TRUE, dev = "pdf",
  fig.height = 5, fig.width = 8, out.width = "\\textwidth"
)
```

@@

<!--
Motivation
PP
Scagnostics
(section needs shortening)
-->
The term "projection pursuit" was coined by @ff74 to describe a procedure for searching high-(say $p$)dimensional data for "interesting" low- dimensional projections ($k=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $k$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $k$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not necessarily found by methods such as principal component analysis.

A large number of projection pursuit indices have been developed [@f87] [@CBC92] [@Lee05] [@AHC02] [@CEM:CEM2568] [@JS87] [@5508437]. Many of these are developed to detect departure from multivariate normal, which includes clusters or outliers, or nonlinearity. Some have been developed to detect separations between known groups. However, if one wants to find complicated co-dependence between parameters there are no options.
On the other hand, for the related question of the selection of pair-wise scatter plots in datasets with large numbers of parameters, a number of different methods have been developed, from detection of linear relations by correlation coefficients, to more complex measures aiming to capture generally the information in a given view.
Here we want to extend the list of projection pursuit indices with the aim of detecting non-trivial dependence between parameter combinations, i.e. we want to find complicated functional dependence as well as association not described by a function (e.g. an ellipse or non-coexistence).
Our motivation is based on potential physics applications aiding the interpretation of model fits to experimental results. Let us consider a physical model with a set of $n$ free parameters that cannot be measured directly but should be determined from fitting a set of $p$ experimental observations for which predictions can be calculated when fixing all $n$ parameters. Note here that while we often have analytic expressions for the predictions, this is not always the case and we may have to rely on numeric calculations. Moreover, a single prediction can be a complicated function of all free parameters. In addition we often deal with large number of observations, $p \sim 100-1000$ as well as sizable parameter spaces $n \sim 10$. The results of a fit are generally interpreted using visualisations selected based on physics intuition and previous knowledge, but the above stated observations suggest that some correlations clearly dictated by the data may be missed, as we are typically only studying selected views of the results. The idea of finding special views in scatter plots fits naturally with the concept of using posterior samples obtained in a Bayesian fit.
The idea is that by finding special views of the posterior samples we gain insight into the physics relevant to accurately describe the experimental data. Apart from improved intuition, finding special regions in parameter space is important as it allows us to further scrutinize the model. There are many examples of this, for example in particle physics detailed studies have been performed e.g. in the alignment limit of the 2HDM or in co-annihilation regions in models of dark matter.

A related question is the selection of pair-wise scatter plots in datasets with large numbers of parameters, for which different methods have been developed. One approach are scatterplot diagnostics, "scagnostics", developed by @scag, and explored in @WW08. \footnote{Scagnostics are available in two R packages @LWscagR and @HWscagR. This work is built upon the scagnostics package by @HWscagR because the underlying base is built using C, works across platforms.} They are calculated from a set of elementary graph-theoretic building blocks: the convex hull, the alpha hull and the minimal spanning tree (MST). These are used to define eight measures: Outlying, Skewed, Sparse, Clumpy, Striated, Convex, Skinny and Stringy. While in principle each measure could be used on its own to define a ranking, the original idea suggested considering pair-wise scatter plots of the scagnostics measures for identification of "characteristic" and "unusual" 2-d views in the original parameter space. Alternative methods suggested by (Katrins Thesis or maybe her R package?) suggest using parallel coordinates, glyphs or a so-called scagramm (based on the idea of corrgrams) that should be coupled to interactive brushing for selection of pair-wise scatter plots to be displayed.
(moved scagnostics definitions to appendix, not sure if needed at all.)
(Issues with scagnostics already for selection of pairs have been described in Katrins Thesis, should have short summary of that here?)

Given the various shortcomings of the scagnostics approach we consider several other measures that can directly be interpreted as ranking of information contained in a given 2-d view.
Two such additional measures have been defined in @mbgraphic. The first is based on smoothing splines, and is defined such that it will take maximum values when the data distribution can be described by a functional form. For a scatterplot with variable X and Y we calculate it as
\begin{equation}
splines_{2d} = max(1-\frac{Var(res_{X\sim Y})}{Var(X)}, 1-\frac{Var(res_{Y\sim X})}{Var(Y)})
\end{equation}
where $Var(.)$ is the variance and $res_{X\sim Y}$ are the residuals in the spline model where $X$ is considered a function of $Y$.
The second new measure introduced in @mbgraphic is aiming to detect also non-functional relations and is based on the idea of distance correlation [@szekely2007]. (need to add definition)
Finally we will also consider the maximal and total information coefficient (MIC and TIC) defined in @Reshef1518 which are based on mutual information found when imposing grids on pair-wise variable plots. These functions are available in R via the interface implemented in the minerva package [@minerva]. 
(read paper and add more detail. how are they similar/different compared to scagnostics and distance correlation? the way that binning is being built into it may make difference in how MIC/TIC may be smoother?)

A projection pursuit index, a function of all possible projections of the data, invariably has many "hills and valleys" and "knife-edge ridges" because of the varying shapes in the underlying density of observations from one projection to the next. From an exploratory data analysis perspective it is interesting to combine numerical optimisation with visualisation to watch the structure of the data moving into a maximum, to jump away from this projection and follow the optimisation again, invariably moving from local maxima to local maxima, and perhaps even a global maximum. Each of these maxima can reveal different information about the data. Projection pursuit optimisation combined with geodesic interpolations between projections provided by a tour [@As85], is called a projection pursuit guided tour [@CBCH94]. Software to run a projection pursuit guided tour is available in the R package, "tourr" [@tourr].
(Motivation strongly relates to intuition, i.e. guided tour, local tour and manual tour may be useful to understand structure in the multivariate context.)
(Even just for the optimisation part the guided tour implementation is useful.)
(Traditional projection pursuit indices are fast to compute and therefore the optimisation would be done in "real time" while displaying the tour. For the measures considered here this generally does not apply, as a consequence we "disentangle" the two into subsequent steps.)

@@

# Constructing a projection pursuit index
\label{sec:construct}

<!--
Definition of pp index
Optimisation
Illustration of what is already available in tourr
New index explanations
-->

@@

<!--
Criteria for a good index
  - smoothness
  - speed
  - fine and broad structure detection
Simulation study
  - example data 
  - sample size, and other factors used
Overview procedures 
  - look at index values of pairs of variables
  - tour betwen pairs of variables
  - optimisation path
-->

@@

((maybe better to start with something like this, though maybe too obvious?))
The various measures defined in the introduction are designed to be computed for all pairs of variables, followed either by selection based on comparing scagnostics scores between all pairs, or by an absolute ranking from most to least interesting (e.g. when using TIC).
Here we modify the usage as follows. The tour implementation is used to generate projections of the multivariate dataset down to two dimensions, i.e. the projected data can be represented as a scatter plot. In this projection the data is represented by two parameters that are defined as combinations of the input parameters, and this parametrisation is passed into the index function calculation to obtain a score for the selected view. Using the guided tour implementation the index functions are optimized such that we find the generalised 2-d parametrisation yielding the most interesting view of the data.

To use an index function for projection pursuit requires that the measure be relatively smooth over the space of all 2D projections, and we will first investigate the smoothness over randomly selected projections. An additional consideration is the computing time, ideally fast computation is preferred as the guided tour can be viewed as the index is being optimized. However, a computationally more costly index may be preferred for better detection properties, and one can consider recording the guided tour for viewing after the optimisation has converged. Finally we note that another important consideration is how the index function changes as the tour moves into the "optimal" view, this will however depend mainly on the distribution rather than the selected index function. (we will see this in the examples)

@@

1. p points on an n-dimensional sphere randomly selected using the geozoo package, as an example of dataset without special views
2. p points where n-1 points are independently drawn from a uniform distribution between [-1,1], and enforcing $x_n^2 + x_{n-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution 
3. p points where n-1 points are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_n = \sin(x_{n-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function

@@

To get an overview of the dataset we show views of the first two parameters showing the independent distribution and the last two parameters showing the special view. The enforced relationships will also affect the 1-d densities as can be seen from \ref{fig:dataPlotsDensity}, which means that views including sizable fractions of V6 (and V5 for the Pipe distribution) will be distinct from uninformed views of distributions 2 and 3. We select two samples for each distribution, with 100 and 1000 sampled points respectively.

```{r dataPlotsV1V2, fig.width=6, fig.height=7, fig.cap="Data views of V1 vs V2."}
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsV5V6, fig.width=6, fig.height=7, fig.cap="Data views of V5 vs V6."}
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V5, V6)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

<!--
its possible to do a single figure legend - need to look up code for this
-->

@@

We next compare the various index values for three views: V1 vs V2, V5 vs V6 and V1 vs V6 (capturing differences from the different distribution in V6). For the Sphere distribution by construction we see only random fluctuations between the different views, which can be sizable in particular for small datasets. Considering next the Pipe distribution, we see that skinny, stringy and dcor2d are increased for the special view, but in particular MIC and TIC index are much larger and may be used to detect this type of structure in the dataset. Moreover, minimising the convex index (or maximising 1-convex) should also allow us to detect the special view. Finally the Sine distribution shows that several of the indices are maximised in the special view, in particular also the splines2d index designed for the detection of fictional dependence.

```{r indexTable, echo=FALSE}
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
indexOverview <- tibble(convex=numeric(),
               clumpy=numeric(),
               skinny=numeric(),
               stringy=numeric(),
               dcor2d=numeric(),
               splines2d=numeric(),
               MIC=numeric(),
               TIC=numeric(),
               name=character(),
               vars=character())
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  ds <- as.tibble(rescale(ds))
  for(vars in list(c("V1","V2"),c("V5","V6"),c("V1","V6"))){
    dprj <- as.matrix(select_(ds, vars[1], vars[2]))
    scagRes <- scagnostics(dprj)
    dcorRes <- dcor2d(dprj[,1], dprj[,2])
    splineRes <- splines2d(dprj[,1], dprj[,2])
    mineRes <- mine(dprj[,1], dprj[,2])
    indexOverview <- add_row(indexOverview, convex=scagRes[,"Convex"],
                    clumpy=scagRes[,"Clumpy"], skinny=scagRes[,"Skinny"],
                    stringy=scagRes[,"Stringy"], dcor2d=dcorRes,
                    splines2d=splineRes, MIC=mineRes$MIC,
                    TIC=mineRes$TIC, name=pLabels[i], vars=paste0(vars[1],"-",vars[2]))
  }
  i <- i+1
}
knitr::kable(indexOverview ,  caption = "Summary of index values.", digits=2) 
```

@@

<!-- index values shouldn't be affected by sample size, on average it should be
same value. Maybe some need to be normalised by n
-->

```{r getProj}
getProj <- function(df, tPath, nameStr, size){
  sc <- tibble(
    holes=numeric(),
    cmass=numeric(),
    convex=numeric(),
    clumpy=numeric(),
    skinny=numeric(),
    stringy=numeric(),
    dcor2d=numeric(),
    splines2d=numeric(),
    MIC=numeric(),
    TIC1=numeric(),
    t=numeric(),
    name=character(),
    size=numeric())
  n <- length(tPath)
  df <- rescale(df)
  for (i in 1:n) {
    dprj <- df %*% tPath[[i]]
    scagRes <- scagnostics(dprj)
    dcorRes <- dcor2d(dprj[,1], dprj[,2])
    splineRes <- splines2d(dprj[,1], dprj[,2])
    mineRes <- mine(dprj[,1], dprj[,2])
    holesRes <- holes(dprj)
    cmassRes <- cmass(dprj)
    sc <- add_row(sc, holes=holesRes, cmass=cmassRes,
                  convex=scagRes[,"Convex"],
                  clumpy=scagRes[,"Clumpy"], skinny=scagRes[,"Skinny"],
                  stringy=scagRes[,"Stringy"], dcor2d=dcorRes,
                  splines2d=splineRes, MIC=mineRes$MIC,
                  TIC1=mineRes$TIC, t=i, name=nameStr, size=size)
  }
  maxTIC <- max(sc$TIC1)
  sc <- sc %>%
    mutate(TIC = TIC1/maxTIC) %>%
    select(-TIC1)

@@

Sharp jumps can be observed in all measures except the distance correlation measure when considering small datasets of only 100 points. Generally we observe smoother behaviour for larger datasets, notably all considered scagnostics measures show only small variation around a flat baseline. Interestingly all other indices indicate preferred viewing directions for dataset 3 when studying large samples, but not for dataset 2 which contains a special view in the same direction. We therefore show 2-d view defined by the tour step $t=58$, with projection matrix

```{r t58matrix, echo=FALSE}
knitr::kable(t1full[[58]] ,  caption = "Matrix representation of projection 58.")
```

i.e. while the last two parameters enter significantly, they will not dominate the view and we do not expect to fully observe the built in features.

```{r plot58, fig.width=6, fig.height=7, fig.cap="View of all datasets defined by at tour step 58."}
p58 <- t1full[[58]]
pL <- list()
i <- 1
pLabels <- c("Sphere 100", "Sphere 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(sphere100, sphere1000, pipe100, pipe1000, sin100, sin1000)){
  dm <- rescale(ds)
  pData <- as.tibble(dm %*% p58)
  pC <- ggplot(pData, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]],
             pL[[3]], pL[[4]],
             pL[[5]], pL[[6]],
             ncol=2)

@@

We see from Fig. ((how to add cross references?)) that the built in special views cannot be seen in this projection, but confirm that the sine 1000 dataset presents some correlation in this view.

## Transition
We repeat the same exercise, but now we are interested in how the index values change when going from the view in V1 vs V2 to the view in V5 vs V6.
```{r plannedtourpath}
m1 <- matrix(c(1,0,0,0,0,0,0,1,0,0,0,0),ncol=2)
m2 <- matrix(c(0,0,0,0,1,0,0,0,0,0,0,1),ncol=2)
#this is silly but seems that first two entries are being ignored, so need some fake entries
m3 <- matrix(c(1,0,0,0,1,0,0,1,0,0,0,1),ncol=2)
m4 <- matrix(c(1,1,1,0,0,0,0,1,1,0,0,0),ncol=2)
t2 <- save_history(sin100,tour_path=planned_tour(list(m3,m4,m1,m2)))
t2full <- as.list(interpolate(t2))
fullResPlanned <- getProj(sin100, t2full, "sin", 100) %>%
  rbind(getProj(sin1000, t2full, "sin", 1000)) %>%
  rbind(getProj(sphere100, t2full, "sphere", 100)) %>%
  rbind(getProj(sphere1000, t2full, "sphere", 1000)) %>%
  rbind(getProj(pipe100, t2full, "pipe", 100)) %>%
  rbind(getProj(pipe1000, t2full, "pipe", 1000))
```

@@

We can see a pretty clear and smooth increase in dcor2d and splines2d as well as MIC and TIC index when approaching the special view of the Sine distribution, which should therefore be efficient to optimise. On the other hand while we also see smooth increase in MIC and TIC when approaching the special view of the Pipe distribution, the increase only becomes apparent much closer to the final projection, making this structure more challenging to detect with a guided tour.
Note that the various indices appear less noisy than for the grand tour shown before, but should note that the tour path here is shorter, and the y-axis scale is different (typically reaching higher index values in this example, whereas smaller fluctuations are more amplified in the previous views).
(What does this mean for optimisation? How can I detect such "smaller scale increase"? Maybe try "search_better" instead of search_geodesic? increase alpha and decrease cooling? increase max.tries?)
Can force it when using search_better with large alpha, cooling=1, max.tries=1000, but not ideal..
Difficulties also from "fake" structure from the sampling boundaries?
Possible issue with search_geodesic is that the search window is fixed in the code..

@@

Here we also see how the existing indices in the tourr package are inadequate for the detection of these type of dependence, the holes index taking maximum values before the final view and decreasing when moving into the special view. On the other hand, surprisingly the cmass index increases when moving into the special view, but the intermediate minimum makes it unsuitable as a projection pursuit index here. QUESTION: how come they are all almost the same for the various distributions? Especially pipe and sphere seem to give almost identical behaviour, can it be understood from the index definition?
